Disentangling the spinal mechanisms of illusory heat and burning sensations in the Thermal Grill Illusion

### 2023-07-17 Manuscript.Rmd
updated_phrasing_of_manuscript_and_organization

---
title: "Disentangling the spinal mechanisms of illusory heat and burning sensations in the Thermal Grill Illusion"
author: 
  - name : "Alexandra G. Mitchell"
    affiliation : "1"
  - name : "Jesper Fischer Ehmsen"
    affiliation : "1"
  - name : "Daniel Elmstrom Christensen"
    affiliation : "2"
  - name : "Anna Villaume Stuckert"
    affiliation : "3"
  - name : "Patrick Haggard"
    affiliation : "4"  
  - name : "Francesca Fardo"
    affiliation : "1"  
affiliation:
  - id : "1"
    institution : "Center of Functionally Integrative Neuroscience (CFIN), Department of Clinical Medicine, Aarhus University, Denmark"
  - id : "2"
    institution : "Interacting Minds Center (IMC), Department XXX, Aarhus University, Denmark"
  - id : "3"
    institution : "Departement of Neuroscience, University of Copenhagen, Denmark"
  - id : "4"
    institution : "Institute of Cognitive Neuroscience, University College London, United Kingdom"
  - id : "5"
    institution : "Danish Pain Research Center, Aarhus University, Denmark"
#date: "`r Sys.Date()`"
output:
  word_document:
  html_document: default
  pdf_document:
    latex_engine: xelatex
bibliography: knitting files/Disentangling the spinal mechanisms of illusory heat and burning sensations in the Thermal Grill Illusion.bib
link-citations: yes
linkcolor: blue
always_allow_html: yes
editor_options: 
  markdown: 
    wrap: sentence
---

@@

```{r, results="hide", fig.width = 7.2, fig.height = 5, warning = F, message = F, echo = F, fig.cap = " Difference in VAS ratings between within and across dermatome conditions for each type of stimulation (Non-TGI and TGI) in each VAS rating quality (cold, warm and burning). Positive values represent higher ratings within dermatomes, negative values represent higher ratings across dermatomes. Small dots are individual means, large dots are population means for each condition and error bars are 95% confidence intervals."}
plot1()
```

In keeping with the unique heat and burning features of TGI, our results exhibited a more robust TGI when stimuli were confined within dermatomes compared to when applied across dermatomes, corresponding to non-adjacent spinal segments (Fig. 2).
When rating the cold thermode (Exp. 1), participants reported a significantly reduced subjective experience of cold for TGI, but not non-TGI, stimuli applied within a dermatome compared to across dermatomes (stimulation by dermatome interaction: $\beta$ = `r stats_model_cold_exp1$beta[6]`, `r make_pvalue(stats_model_cold_exp1$p[6])`).
However we did not observed such interaction effect for warm ($\beta$ = `r stats_model_warm_exp1$beta[6]`, `r make_pvalue(stats_model_warm_exp1$p[6])`) and burning ratings ($\beta$ = `r stats_model_burn_exp1$beta[6]`, `r make_pvalue(stats_model_burn_exp1$p[6])`).

In contrast, when rating the warm thermode (Exp. 2), participants reported markedly enhanced burning sensations for TGI, but not non-TGI, stimuli applied within a dermatome compared to across dermatomes (stimulation by dermatome interaction: $\beta$ = `r stats_model_burn_exp2$beta[6]`, `r make_pvalue(stats_model_burn_exp2$p[6])`).
However, we did not observe a modulation of cold ($\beta$ = `r stats_model_cold_exp2$beta[6]`, `r make_pvalue(stats_model_cold_exp2$p[6])`) ) and warm thermosensory ratings ($\beta$ = `r stats_model_warm_exp2$beta[6]`, `r make_pvalue(stats_model_warm_exp2$p[6])`).

@@

Previous research demonstrated a phenomenon known as distal inhibition, wherein heat pain ratings tend to increase when a participant evaluates a more distal compared to a more proximal stimulus among two warm stimuli presented on the forearm [@quevedo_psychophysicshyperalgesia_2004; @quevedo_illusion_2007].
In our study, we sought to investigate whether this effect of distal inhibition would also influence TGI perception within a single dermatome (Fig. 3).
We tested this effect with a two way interaction between condition (control / TGI) and the location of the cold thermode when it was placed within dermatome (distal / proximal) Results showed no statistical significant interaction effect for the cold ratings in both experiments: (Exp. 1: $\beta$ = `r stats_model_cold_exp1$beta[8]`, `r make_pvalue(stats_model_cold_exp1$p[8])`; Exp. 2: $\beta$ = `r stats_model_cold_exp2$beta[8]`, `r make_pvalue(stats_model_cold_exp2$p[8])`) Neither a statistically significant effect for warm watings (Exp1: $\beta$ = `r stats_model_warm_exp1$beta[8]`, `r make_pvalue(stats_model_warm_exp1$p[8])`; Exp 2: $\beta$ = `r stats_model_warm_exp2$beta[8]`, `r make_pvalue(stats_model_warm_exp2$p[8])`) and also not for burning ratings (Exp1: $\beta$ = `r stats_model_burn_exp1$beta[8]`, `r make_pvalue(stats_model_burn_exp1$p[8])`; Exp 2: $\beta$ = `r stats_model_burn_exp2$beta[8]`, `r make_pvalue(stats_model_burn_exp2$p[8])`)

```{r, results="hide", fig.width = 7.2, fig.height = 7.2, warning = F, message = F, echo = F, fig.cap = "Difference in VAS ratings between the two dermatome conditions. The top panel shows the difference between proximal and distal cold afferent locations (within dermatomes) and the bottom panel shows the difference between rostral and caudal cold afferent locations, relative to warm, across stimulation type (Non-TGI, TGI) and VAS rating quality (cold, warm and burning). Small points show data from each participant, large dots are means across trials for each condition, and error bars show 95% confidence intervals."}
plot2()
```

@@

A main objective of this study was assessing spatial order effects along the rostrocaudal axis at the spinal level.
We delivered an equal number of trials in which the cold stimulus was applied on a dermatome that mapped more rostrally or caudally compared to the warm or neutral stimuli.
Like the previous analysis we investigated this relationship with a two way interaction here with the rostral - caudal axis being investigated.
Results showed no statistical significant interaction effect for the cold ratings in the first experiment, but a statistically significant interaction when the reference thermode was warm: (Exp. 1: $\beta$ = `r stats_model_cold_exp1$beta[7]`, `r make_pvalue(stats_model_cold_exp1$p[7])`; Exp. 2: $\beta$ = `r stats_model_cold_exp2$beta[7]`, `r make_pvalue(stats_model_cold_exp2$p[7])`) No statistically significant effect for warm watings (Exp1: $\beta$ = `r stats_model_warm_exp1$beta[7]`, `r make_pvalue(stats_model_warm_exp1$p[7])`; Exp 2: $\beta$ = `r stats_model_warm_exp2$beta[7]`, `r make_pvalue(stats_model_warm_exp2$p[7])`) and also not for burning ratings (Exp1: $\beta$ = `r stats_model_burn_exp1$beta[7]`, `r make_pvalue(stats_model_burn_exp1$p[7])`; Exp 2: $\beta$ = `r stats_model_burn_exp2$beta[7]`, `r make_pvalue(stats_model_burn_exp2$p[7])`)

>>>>

---
title: "Disentangling the spinal mechanisms of illusory heat and burning sensations in the Thermal Grill Illusion"
author: 
  - name : "Alexandra G. Mitchell"
    affiliation : "1"
  - name : "Jesper Fischer Ehmsen"
    affiliation : "1"
  - name : "Daniel Elmstrom Christensen"
    affiliation : "2"
  - name : "Anna Villaume Stuckert"
    affiliation : "3"
  - name : "Patrick Haggard"
    affiliation : "4"  
  - name : "Francesca Fardo"
    affiliation : "1"  
affiliation:
  - id : "1"
    institution : "Center of Functionally Integrative Neuroscience (CFIN), Department of Clinical Medicine, Aarhus University, Denmark"
  - id : "2"
    institution : "Interacting Minds Center (IMC), Department XXX, Aarhus University, Denmark"
  - id : "3"
    institution : "Departement of Neuroscience, University of Copenhagen, Denmark"
  - id : "4"
    institution : "Institute of Cognitive Neuroscience, University College London, United Kingdom"
  - id : "5"
    institution : "Danish Pain Research Center, Aarhus University, Denmark"
#date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document:
    latex_engine: xelatex
  word_document:
bibliography: knitting files/Disentangling the spinal mechanisms of illusory heat and burning sensations in the Thermal Grill Illusion.bib
link-citations: yes
linkcolor: blue
always_allow_html: yes
editor_options: 
  markdown: 
    wrap: sentence
---

@@

```{r, results="hide", fig.width = 7.2, fig.height = 5, warning = F, message = F, echo = F, fig.cap = " Difference in VAS ratings between within and across dermatome conditions for each type of stimulation (Non-TGI and TGI) in each VAS rating quality (cold, warm and burning). Positive values represent higher ratings within dermatomes, negative values represent higher ratings across dermatomes. Small dots are individual means, large dots are population means for each condition and error bars are 95% confidence intervals."}

plot1(include_zero = F)

#plot1(include_zero = T)

In keeping with the unique heat and burning features of TGI, our results exhibited a more robust TGI when stimuli were confined within dermatomes compared to when applied across dermatomes, corresponding to non-adjacent spinal segments (Fig. 2).

When rating the cold thermode (Exp. 1), participants reported a significantly reduced subjective experience of cold for TGI, but not non-TGI, stimuli applied within a dermatome compared to across dermatomes (stimulation by dermatome interaction: $\beta$ = `r stats_model_cold_exp1$beta[6]`, `r make_pvalue(stats_model_cold_exp1$p[6])`). This interaction was not present for for warm ($\beta$ = `r stats_model_warm_exp1$beta[6]`, `r make_pvalue(stats_model_warm_exp1$p[6])`), but we did find a strong main effect of dermatome ($\beta$ = `r stats_model_warm_exp1$beta[2]`, `r make_pvalue(stats_model_warm_exp1$p[2])`) For the burning ratings we found No significant interaction  ($\beta$ = `r stats_model_burn_exp1$beta[6]`, `r make_pvalue(stats_model_burn_exp1$p[6])`) nor main effect ($\beta$ = `r stats_model_burn_exp1$beta[2]`, `r make_pvalue(stats_model_burn_exp1$p[2])`).

In contrast, when rating the warm thermode, participants reported markedly enhanced burning sensations for TGI, but not non-TGI, (stimulation by dermatome interaction: $\beta$ = `r stats_model_burn_exp2$beta[6]`, `r make_pvalue(stats_model_burn_exp2$p[6])`).However, we did not observe suvh a modulation of cold percepts ($\beta$ = `r stats_model_cold_exp2$beta[6]`, `r make_pvalue(stats_model_cold_exp2$p[6])`), with no main effect ($\beta$ = `r stats_model_cold_exp2$beta[2]`, `r make_pvalue(stats_model_cold_exp2$p[2])`) for the warm thermosensory ratings we found no interaction ($\beta$ = `r stats_model_warm_exp2$beta[6]`, `r make_pvalue(stats_model_warm_exp2$p[6])`), and no main effect of dermatome ($\beta$ = `r stats_model_warm_exp2$beta[2]`, `r make_pvalue(stats_model_warm_exp2$p[2])`).

@@

Previous research demonstrated a phenomenon known as distal inhibition, wherein heat pain ratings tend to increase when a participant evaluates a more distal compared to a more proximal stimulus among two warm stimuli presented on the forearm [@quevedo_psychophysicshyperalgesia_2004; @quevedo_illusion_2007].
In our study, we sought to investigate whether this effect of distal inhibition would also influence TGI perception within a single dermatome (Fig. 3).
We tested this effect with a two way interaction between condition (control / TGI) and the location of the cold thermode when it was placed within dermatome (distal / proximal). In the first experiment with the cold thermode as the reference we found no statistical significant interaction effect for the cold ratings ($\beta$ = `r stats_model_cold_exp1$beta[8]`, `r make_pvalue(stats_model_cold_exp1$p[8])`, however with a strong main effect of dermatome ($\beta$ = `r stats_model_cold_exp1$beta[4]`, `r make_pvalue(stats_model_cold_exp1$p[4])`) For the warm ratings we found no interaction ($\beta$ = `r stats_model_warm_exp1$beta[8]`, `r make_pvalue(stats_model_warm_exp1$p[8])`), nor main effect of dermatome ($\beta$ = `r stats_model_warm_exp1$beta[4]`, `r make_pvalue(stats_model_warm_exp1$p[4])`)
For the burning ratings We found no significant interaction ($\beta$ = `r stats_model_burn_exp1$beta[8]`, `r make_pvalue(stats_model_burn_exp1$p[8])`), but a significant main effect ( $\beta$ = `r stats_model_burn_exp1$beta[4]`, `r make_pvalue(stats_model_burn_exp1$p[4])`)

For the second experiment with the warm thermode as the reference we found no interaction no interaction ($\beta$ = `r stats_model_burn_exp2$beta[8]`, `r make_pvalue(stats_model_burn_exp2$p[8])`), nor main effect ($\beta$ = `r stats_model_burn_exp2$beta[4]`, `r make_pvalue(stats_model_burn_exp2$p[4])`) for the burning ratings.
For the warm ratings we found no interaction ($\beta$ = `r stats_model_warm_exp2$beta[8]`, `r make_pvalue(stats_model_warm_exp2$p[8])`), , nor main effect of dermatome ($\beta$ = `r stats_model_warm_exp2$beta[4]`, `r make_pvalue(stats_model_warm_exp2$p[4])`) and for the cold ratings no interaction ($\beta$ = `r stats_model_cold_exp2$beta[8]`, `r make_pvalue(stats_model_cold_exp2$p[8])`) but a main effect ($\beta$ = `r stats_model_cold_exp2$beta[4]`, `r make_pvalue(stats_model_cold_exp2$p[4])`).

```{r, results="hide", fig.width = 7.2, fig.height = 7.2, warning = F, message = F, echo = F, fig.cap = "Difference in VAS ratings between the two dermatome conditions. The top panel shows the difference between proximal and distal cold afferent locations (within dermatomes) and the bottom panel shows the difference between rostral and caudal cold afferent locations, relative to warm, across stimulation type (Non-TGI, TGI) and VAS rating quality (cold, warm and burning). Small points show data from each participant, large dots are means across trials for each condition, and error bars show 95% confidence intervals."}

plot2(include_zero = F)

#plot2(include_zero = T)

@@

A main objective of this study was assessing spatial order effects along the rostrocaudal axis at the spinal level.
We delivered an equal number of trials in which the cold stimulus was applied on a dermatome that mapped more rostrally or caudally compared to the warm or neutral stimuli.
Like the previous analysis we investigated this relationship with a two way interaction here with the rostral - caudal axis being investigated.

Results showed no statistical significant interaction effect for the cold ratings in the first experiment with the reference thermode being cold ($\beta$ = `r stats_model_cold_exp1$beta[7]`, `r make_pvalue(stats_model_cold_exp1$p[7])`, but a main effect ($\beta$ = `r stats_model_cold_exp1$beta[3]`, `r make_pvalue(stats_model_cold_exp1$p[3])`).
For the warm ratings we found no statistically significant effect for warm watings ($\beta$ = `r stats_model_warm_exp1$beta[7]`, `r make_pvalue(stats_model_warm_exp1$p[7])`), but a statistical significant main effect ($\beta$ = `r stats_model_warm_exp1$beta[3]`, `r make_pvalue(stats_model_warm_exp1$p[3])`)
The burning ratings showed no significant interaction ($\beta$ = `r stats_model_burn_exp1$beta[7]`, `r make_pvalue(stats_model_burn_exp1$p[7])`), nor main effect ($\beta$ = `r stats_model_burn_exp1$beta[3]`, `r make_pvalue(stats_model_burn_exp1$p[3])`).

In Experiment 2, with the warm thermode as the reference we found no significant interaction effect for the burning ratings ($\beta$ = `r stats_model_burn_exp2$beta[7]`, `r make_pvalue(stats_model_burn_exp2$p[7])`), but a significant main effect ($\beta$ = `r stats_model_burn_exp2$beta[3]`, `r make_pvalue(stats_model_burn_exp2$p[3])`)
For the warm ratings we found no statistically significant interaction ($\beta$ = `r stats_model_warm_exp2$beta[7]`, `r make_pvalue(stats_model_warm_exp2$p[7])`), but a strong main effect ($\beta$ = `r stats_model_warm_exp2$beta[3]`, `r make_pvalue(stats_model_warm_exp2$p[3])`)
For the cold ratings we found a significant interaction ($\beta$ = `r stats_model_cold_exp2$beta[7]`, `r make_pvalue(stats_model_cold_exp2$p[7])`)
